---
title: "Heatmap IRF4 bound"
output:
  html_document:
    df_print: paged
---

The goal of this analysis is to recreate one analysis that Fede performed that generated a heatmap that was highly particular.
It featured two datasets we had generated as well as a dataset that contained the genes bound by IRF4 at specific timepoints in B cells (Sciamas et al.).
The logical behind the heatmap can be explained as:

1) genes were included into the analysis, that showed > 2-fold expression change between Irf4–/– B220+ mIgM- pro/preB cells and leukemia samples
2) genes were filtered to only include genes that are also differentially regulated in IRF4-RV or EV-RV transduced T11 cells
3) resultant genes were limited to genes previously shown to be bound by IRF4 using ChIP-seq (Sciamas et al.)

### 1) step 
```{r}
library(tidyverse)
library(readxl)
TumBM <- read_excel("TumBM.xlsx")
IRF4GFP <- read_excel("IRF4vsGFP.xlsx")

# import the TumBM scenario and create a column of Log2(foldchange)
# then filtering only the genes to which the criterium of step 1 applies
TumBM_red_1 <- TumBM %>%
  mutate( 
    FoldChange = (TUM1 + TUM3) / 2 / ((BM1 + BM2) / 2)
    ) %>%
  mutate(
    log2FoldChange = log2(FoldChange)
    ) %>%
  rename(
    "geneSymbol" = "NAME"
    ) %>%
  filter(
    (log2FoldChange >= 1) & (log2FoldChange <= 100) |
    ((log2FoldChange <= -1) & (log2FoldChange >= -100))
    ) %>%
  select(-DESCRIPTION, -FoldChange)

# creating a vector of the genes that remain as a reference for filtering
Genes_interest_1 <- TumBM_red_1$geneSymbol %>%
  unlist()

# filtering the IRF4GFP scenario with these genes
IRF4GFP_red_1 <- IRF4GFP %>%
  select(geneSymbol, log2FoldChange, GFP1:IRF42) %>%
  filter(
    geneSymbol %in% Genes_interest_1
  )
```

### 2) step
```{r}
# searching for the genes that fullfil criterium 2
IRF4GFP_red_2 <- IRF4GFP_red_1 %>%
   mutate( 
    FoldChange = (IRF41 + IRF42) / 2 / ((GFP1 + GFP2) / 2)
    ) %>%
  mutate(
    log2FoldChange = log2(FoldChange)
    ) %>%
  filter(
    (log2FoldChange <= 0) |
    (log2FoldChange >= 0.2)
  ) %>%
  select(-FoldChange)


# creating a vector that contains all the genes that are of interest
Genes_interest_2 <- IRF4GFP_red_2$geneSymbol %>%
  unlist()

# the filtering of the TUMBM scenario
TumBM_red_2 <- TumBM_red_1 %>%
  filter(
    geneSymbol %in% Genes_interest_2
  )
```

### 3) step (the lazy version)
Because I trust Fede I will use the genes he found to remain instead of the whole Sciamas dataset
```{r}
# creating a vector of the genes of interest
Genes_interest_3 <- c("Plxna2", "Prkch", "Src", "Ptk2", "Tox", "Mybbp1a",
                      "Jdp2", "Rbpms", "Noa1", "Slc25a15", "Phgdh", "Rps14",
                      "Ppa2", "Gemin6", "Txndc11", "Bach2", "Tram2", "Mpeg",
                      "Nav2", "Cacnae1"
                      )

# filtering both datatables with the third criterium
IRF4GFP_red_3 <- IRF4GFP_red_2 %>%
  filter(
    geneSymbol %in% Genes_interest_3
  ) %>%
  rename(
    "log2FoldChange_IRF4" = "log2FoldChange"
  )

TumBM_red_3 <- TumBM_red_2 %>%
  filter(
    geneSymbol %in% Genes_interest_3
  ) %>%
  rename(
    "log2FoldChange_TUM" = "log2FoldChange"
  )

# combining the two tibbles
combined <- left_join(IRF4GFP_red_3, TumBM_red_3)

```

### Plotting the Heatmap
```{r}
install.packages("dendsort")
library(dendsort)
library(pheatmap)
library(svglite)

# Step 1: Create a row-names vector for plotting
combined <- combined %>%
  arrange(log2FoldChange_TUM)

Row_names <- as_vector(combined$geneSymbol)

# Step 2: select the columns to plot
combined_plot <- combined %>%
  select(
    log2FoldChange_IRF4,
    log2FoldChange_TUM
  ) %>%
  relocate(
    log2FoldChange_TUM
  )

# Step 3: plot using pheatmap
dat <- as.matrix(combined_plot)
rownames(dat) <- Row_names
plot <- pheatmap(dat,
         scale = "column",
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         gaps_row = 4,
         fontsize_row = 6,
         cutree_rows = 2
)
    
# Step 4: write a function for export and export
save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}

save_pheatmap_pdf(plot, "plot.pdf")

# Clustering of the dendogram sorted -> is not used
sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...)))
dat_cluster_rows <- sort_hclust(hclust(dist(dat)))

```

